---
title: "Applications"
description: "Downloadable data and step-by-step replication workflows for two SpatialEffect2 empirical applications."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This page provides step-by-step replication entry points for the two applications in `WangSamiiChangAronow_2024_replication/application`. Each application includes a one-click download bundle (data + script) and a guided workflow.

## 1. Ferraro et al. (2015): Costa Rica Protected Areas

<a class="btn btn-primary" href="../applications/ferraro_etal_2015_sf_bundle.zip" download>Download Ferraro Data and Script (.zip)</a>

### Step 1. Unzip and Keep the Directory Structure

Please unzip to the following structure:

```text
Ferraro_etal_2015_sf/
  data/
  scripts/Ferraro_etal_2015_sf.R
```

### Step 2. Install Dependencies

```{r}
install.packages(c("SpatialEffect2", "sf", "terra", "fields"))
```

### Step 3. Run the Replication Script

```{r}
source("Ferraro_etal_2015_sf/scripts/Ferraro_etal_2015_sf.R")
```

### Step 4. Check the Outputs

The script will generate:

- Main result object: `data/Ferraro_etal_2015_sf.RData`
- Figures: `graphs/` (including AME curves and propensity score plots)


## 2. Jayachandran et al. (2017): Uganda RCT

<a class="btn btn-primary" href="../applications/jayachandran_etal_2017_bundle.zip" download>Download Jayachandran Data and Script (.zip)</a>

### Step 1. Unzip and Keep the Directory Structure

Please unzip to the following structure:

```text
Jayachandran_etal_2017/
  data/
  scripts/Jayachandran_etal_2017.R
```

### Step 2. Install Dependencies

```{r}
install.packages(c("SpatialEffect2", "sf", "sp", "raster", "terra", "fields", "ggplot2"))
```

### Step 3. Run the Replication Script

```{r}
source("Jayachandran_etal_2017/scripts/Jayachandran_etal_2017.R")
```

### Step 4. Check the Outputs

The script will generate:

- Main result object: `data/forestResult_reproj.RData`
- Figures: `graphs/` and `graphs/J2017_graphs/`


## Notes

- Both scripts set a fixed random seed (`set.seed(2024)`) for reproducibility.
- Runtime depends on machine performance; randomization inference and spatial HAC standard error computation can take noticeable time.
