---
title: "Ferraro et al. (2015)"
description: "A concise Costa Rica protected-area application for SpatialEffect2."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ferraro et al. (2015)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  dev.args = list(bg = "white")
)

locate_bundle <- function(bundle_name) {
  candidates <- c(
    file.path("pkgdown", "assets", "applications", bundle_name),
    file.path("..", "pkgdown", "assets", "applications", bundle_name),
    file.path("docs", "applications", bundle_name),
    file.path("..", "docs", "applications", bundle_name)
  )
  hit <- candidates[file.exists(candidates)]
  if (length(hit) == 0) return(NA_character_)
  hit[[1]]
}

prepare_ferraro_inputs <- function(protected_areas, forest,
                                   buffer_distance = 5000,
                                   coarse_res = 5000,
                                   fine_res = 2000,
                                   max_distance = 20000) {
  crs_pa <- sf::st_crs(protected_areas)
  # Use proj4string for terra::rast(); the shapefile WKT contains Latin-1
  # characters that break tolower() under UTF-8 locales (e.g. Linux CI).
  crs_arg <- crs_pa$proj4string
  if (is.na(crs_arg) || !nzchar(crs_arg)) {
    crs_arg <- paste0("EPSG:", crs_pa$epsg)
  }

  pa_buffers <- sf::st_buffer(protected_areas, dist = buffer_distance)

  bbox <- sf::st_bbox(pa_buffers)
  blank_rast <- terra::rast(
    terra::ext(bbox["xmin"], bbox["xmax"], bbox["ymin"], bbox["ymax"]),
    res = coarse_res,
    crs = crs_arg
  )

  pa_rast <- terra::rasterize(
    terra::vect(protected_areas), blank_rast,
    field = 1, fun = "mean", background = 0
  )
  buf_rast <- terra::rasterize(
    terra::vect(pa_buffers), blank_rast,
    field = 1, fun = "mean", background = 0
  )

  pa_rast[pa_rast == 0] <- NA
  buf_rast[buf_rast == 0] <- NA
  buf_rast[buf_rast == 1] <- 0
  combined_rast <- terra::cover(pa_rast, buf_rast)

  boundary_rast <- terra::focal(
    combined_rast,
    w = matrix(1, 3, 3),
    fun = function(x) {
      ux <- unique(x[!is.na(x)])
      if (length(ux) > 1) 1 else NA
    },
    na.policy = "omit"
  )
  boundary_rast[is.na(combined_rast)] <- NA
  boundary_rast[combined_rast == 0] <- 0

  analysis_area <- sf::st_buffer(pa_buffers, dist = max_distance + fine_res)
  analysis_bbox <- sf::st_bbox(analysis_area)
  rast_template <- terra::rast(
    terra::ext(
      analysis_bbox["xmin"], analysis_bbox["xmax"],
      analysis_bbox["ymin"], analysis_bbox["ymax"]
    ),
    res = fine_res,
    crs = crs_arg
  )

  ras <- terra::rasterize(
    terra::vect(forest), rast_template,
    field = "defor97", fun = mean, na.rm = TRUE
  )
  ras[is.na(ras)] <- 0
  names(ras) <- "defor97"

  forest$luc_high <- as.numeric(forest$luc1 == 1 | forest$luc2 == 1 | forest$luc3 == 1)
  ras_cov1 <- terra::rasterize(terra::vect(forest), rast_template, field = "luc_high", fun = mean, na.rm = TRUE)
  ras_cov2 <- terra::rasterize(terra::vect(forest), rast_template, field = "dist_rd_69", fun = mean, na.rm = TRUE)
  ras_cov3 <- terra::rasterize(terra::vect(forest), rast_template, field = "dist_mcity", fun = mean, na.rm = TRUE)

  ras_cov1[is.na(ras_cov1)] <- 0
  ras_cov2[is.na(ras_cov2)] <- 0
  ras_cov3[is.na(ras_cov3)] <- 0
  names(ras_cov1) <- "luc_high"
  names(ras_cov2) <- "dist_rd_69"
  names(ras_cov3) <- "dist_mcity"

  ras_Z_sf <- sf::st_as_sf(terra::as.polygons(boundary_rast, aggregate = FALSE))
  value_cols <- setdiff(names(ras_Z_sf), attr(ras_Z_sf, "sf_column"))
  treatment_col <- value_cols[1]

  node_means <- terra::extract(
    c(ras, ras_cov1, ras_cov2, ras_cov3),
    terra::vect(ras_Z_sf),
    fun = mean,
    na.rm = TRUE
  )

  Zdata <- data.frame(
    treatment = as.integer(ras_Z_sf[[treatment_col]]),
    luc_high = node_means$luc_high,
    dist_road = node_means$dist_rd_69,
    dist_city = node_means$dist_mcity
  )

  keep <- is.finite(Zdata$treatment) &
    is.finite(Zdata$luc_high) &
    is.finite(Zdata$dist_road) &
    is.finite(Zdata$dist_city)

  list(
    ras = ras,
    ras_Z_sf = ras_Z_sf[keep, , drop = FALSE],
    Zdata = Zdata[keep, , drop = FALSE],
    boundary_counts = table(terra::values(boundary_rast)[, 1], useNA = "ifany")
  )
}
```

## 1. Background

This example follows Ferraro et al. (2015), an observational study of protected
areas and forest outcomes in Costa Rica. Protected areas are not randomly placed:
they may differ from nearby unprotected land in soil, road access, city access,
and other conditions. We use SpatialEffect2 to estimate how deforestation differs
near protected-area boundaries after a simple propensity-score adjustment.

<a class="btn btn-primary" href="../applications/ferraro_etal_2015_sf_bundle.zip" download>Download Ferraro Data Bundle (.zip)</a>

## 2. Load Data

```{r}
library(SpatialEffect2)
library(terra)
library(sf)

ferraro_bundle <- locate_bundle("ferraro_etal_2015_sf_bundle.zip")
stopifnot(!is.na(ferraro_bundle))

td <- tempfile("ferraro2015_")
dir.create(td)
unzip(ferraro_bundle, exdir = td)

base_dir <- file.path(td, "ferraro_etal_2015_sf")
pa_path <- file.path(base_dir, "data", "Costa Rica", "PAs_Reproj", "PAs_before80_1.shp")
forest_path <- file.path(
  base_dir,
  "data", "Costa Rica", "Deforestation and Carbon", "Spatial",
  "final_pixels_w_covs.shp"
)

protected_areas <- st_read(pa_path, quiet = TRUE)
forest <- st_read(forest_path, quiet = TRUE)
forest <- st_transform(forest, st_crs(protected_areas))
protected_areas <- st_make_valid(protected_areas)
```

## 3. Prepare the Analysis Objects

```{r}
# The helper prepares three objects used by SpatialEffect():
# ras: outcome raster; ras_Z_sf: intervention tiles; Zdata: treatment data.
prepared <- prepare_ferraro_inputs(protected_areas, forest)

ras <- prepared$ras
ras_Z_sf <- prepared$ras_Z_sf
Zdata <- prepared$Zdata

c(
  n_protected_areas = nrow(protected_areas),
  n_outcome_parcels = nrow(forest),
  n_nodes = nrow(Zdata),
  n_treated = sum(Zdata$treatment == 1),
  n_control = sum(Zdata$treatment == 0)
)

prepared$boundary_counts
```

The full bundle contains the longer data-preparation script. The website keeps
only the main steps visible so the empirical workflow is easier to follow.

## 4. Visualize the Data

```{r, fig.width=7.4, fig.height=5.8, fig.alt="Deforestation raster overlaid with protected and control boundary tiles."}
op <- par(mar = c(3.3, 3.3, 3.2, 5.2))
plot(
  ras,
  main = "Costa Rica Deforestation and Boundary Tiles",
  col = hcl.colors(60, "YlGnBu", rev = TRUE),
  plg = list(title = "Defor.", cex = 0.8)
)
plot(vect(ras_Z_sf[Zdata$treatment == 0, ]), add = TRUE, border = "#2C7FB8", lwd = 1.0)
plot(vect(ras_Z_sf[Zdata$treatment == 1, ]), add = TRUE, border = "#D94841", lwd = 1.2)
legend(
  "bottomleft",
  legend = c("control", "protected"),
  col = c("#2C7FB8", "#D94841"),
  lwd = 2,
  bg = grDevices::adjustcolor("white", alpha.f = 0.85),
  cex = 0.9
)
par(op)
```

## 5. Propensity Scores and Estimand

```{r}
# Protected-area placement is observational, so we estimate a simple propensity
# score from node-level land and access covariates.
pscore_fit <- glm(
  treatment ~ luc_high + dist_road + dist_city,
  data = Zdata,
  family = binomial()
)

Zdata$prob_treatment <- pmin(pmax(predict(pscore_fit, type = "response"), 0.01), 0.99)
summary(Zdata$prob_treatment)
```

```{r, fig.width=7.4, fig.height=4.8, fig.alt="Density plot of estimated propensity scores for protected-area boundary tiles."}
plot(
  density(Zdata$prob_treatment),
  main = "Estimated Propensity Scores",
  xlab = "Estimated probability of being protected",
  lwd = 2,
  col = "#2C7FB8"
)
```

We target the **Average Marginalized Effect (AME)** at distance `d`: the average
change in deforestation at distance `d` when a boundary tile is protected rather
than unprotected, averaging over the rest of the spatial layout. A negative AME
means less deforestation near protected-area tiles than near comparable control
tiles.

## 6. Estimate with SpatialEffect2

```{r}
dVec_km <- seq(0, 20, by = 2)
dVec_m <- dVec_km * 1000

result <- SpatialEffect(
  ras = ras,
  ras_Z = ras_Z_sf,
  Zdata = Zdata,
  treatment = "treatment",
  prob_treatment = "prob_treatment",
  dVec = dVec_m,
  cType = "donut",
  dist.metric = "Euclidean",
  numpts = 600,
  smooth = 1,
  bw = 5000,
  bw_debias = 5000,
  smooth.conley.se = 0,
  per.se = 0,
  conley.se = 1,
  cutoff = 15000,
  kernel = "tri",
  nPerms = 100,
  n_threads = 1L,
  perm_engine = "shuffle"
)

head(result$AME_est, 6)
```

## 7. Look at the Estimated Results

The returned object has class `"SpatialEffect"`, so `graphics::plot(result, ...)`
uses SpatialEffect2's S3 method, `plot.SpatialEffect()`. Because `conley.se = 1`,
the plot includes Conley confidence intervals.

```{r, fig.width=7.4, fig.height=5.2, fig.alt="Estimated AME curve by distance for the Ferraro protected-area application using the SpatialEffect S3 plot method."}
graphics::plot(
  result,
  smooth = TRUE,
  ci.type = "conley",
  style = "lines",
  xlab = "Distance from protected-area boundary (meters)",
  ylab = "Estimated AME",
  main = "Estimated AME Curve"
)
```

```{r}
summary(result, dVec.range = c(0, 6000))

ame <- result$AME_est
ame$km <- ame$d / 1000
subset(ame, km %in% c(0, 2, 4, 6, 10, 20))
```

## 8. Test a Policy-Relevant Distance Range

```{r}
test <- SpatialEffectTest(
  result.list = result,
  test.range = c(0, 5000),
  smooth = 0,
  alpha = 0.05
)
test

test$test.stat < test$test.CI[1] || test$test.stat > test$test.CI[2]
```

`TRUE` means the observed range-level effect lies outside the permutation null
interval at level `alpha`.

## 9. Interpretation

In this lightweight run, the estimated AME is negative near protected-area
boundaries and moves toward zero as distance increases. Substantively, this
suggests that the strongest association between protected areas and lower
deforestation is local, with weaker differences farther from the boundary.

For a full analysis, use a finer raster, more distance points, and more
permutations. Set `per.se = 1` if you also want permutation intervals at each
distance.

## 10. Reference

Ferraro, P. J., M. M. Hanauer, D. A. Miteva, J. L. Nelson, S. K. Pattanayak,
C. Nolte, and K. R. E. Sims (2015). "Estimating the impacts of conservation on
ecosystem services and poverty by integrating modeling and evaluation."
*Proceedings of the National Academy of Sciences of the United States of
America* 112(24): 7420-7425.
